A Proposed Translation of B Implementations to the LLVM

### 2013-07-20 PAPERS/2013-SBMF/paper_short.tex
no message

LLVM aims to be a general, language agnostic, intermediate representation for
compilers. The compiler front-end translates the source programming language to
LLVM source code, upon which optimization and other static analysis tasks may be
performed, and the compiler back-end translates to the target platform assembly
language. One key feature of LLVM is that it is a single-static assignment (SSA)
language: a variable may only be assigned once. So for each assignment to a
variable found in the source language, the front-end shall generate a new
temporary variable in the LLVM program. For illustration,
figure~\ref{fig:ex-llvm} presents a LLVM program (for clarity, the corresponding
C code appears on the left) with two temporary variables \llvm{\%0} and
\llvm{\%1}.
\begin{figure}
\vspace*{-2.4em}
  \begin{minipage}[t]{.3\textwidth}
    \begin{tabbing}
      foo \= \kill  \\
%      \B{VARIABLES}\\
%      \> y\\
      inc(x) =\\
      \B{PRE} x : INT \B{THEN}   \\
      \> 	y = x+y\\
      \B{END} \\
      \end{tabbing}
  \end{minipage}
  \:\vline\:
  \begin{minipage}[t]{.2\textwidth}
    \begin{tabbing}
      foo \= \kill  \\
%	 int y;\\      
      void inc(int x) \{\\
      \> 	y = x+y; \\
      \}\end{tabbing}
  \end{minipage}
	  \:\vline\:
  \begin{minipage}[t]{.6\textwidth}
    \begin{tabbing}
      foo \= \kill  \\
%	 \llvm{@y = global i32 0 } \\      
      \llvm{define void @inc(i32 \%x) nounwind \{} \\
      \llvm{entry:}                      \\
      \> \llvm{\%0 = load i32* @y, align 4}     \\
      \> \llvm{\%1 = add nsw i32 \%0, \%x}     \\
      \> \llvm{store i32 \%1, i32* @y, align 4} \\
      \> \llvm{ret void} \\
      \llvm{\}}
    \end{tabbing}

@@

\section{Translation: Expressions
  \label{sec:expr}}

To evaluate a composed expression using the LLVM, one must first generate a
sequence of instructions to evaluate each argument, the resulting value being
stored in a temporary variable, and second generate an instance of the LLVM
instruction corresponding to the expression argument.  Translation of
expressions is defined as functions that take as input some class of expressions
and produce a triple of strings $p \sep v \sep t$, where $p$ is a possibly empty
sequence of instructions required in the evaluation (e.g.  for the
sub-expressions), $v$ is the LLVM expression representing the resulting value,
and $t$ represents the LLVM type of the expression. In the case of composed
predicates, though, the evaluation is different as boolean operations need to be
implemented with conditional and unconditional jumps.

\paragraph{Atomic expressions} have a simple translation. In the case of
literals, one only needs to issue the corresponding LLVM literal:
$\trad{TRUE}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"1"} \sep \llvm{"i1"}$,
$\trad{FALSE}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"0"} \sep \llvm{"i1"}$ and
$\trad{l}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"\PH{l}"} \sep \llvm{"i32"}$ when
$\B{l.type} = \B{INT}$.

For identifiers, three cases are possible: global constants, local variables,
and module state variables. For state variables, the translation issues
statements to assign the address of the corresponding element in the state
structure to pointer variable $p$ and then dereference $p$ to assign $v$:
\begin{align*}
\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
& \IF \B{n} \mbox{ is a constant } \THEN \llvm{""} \sep \Global(\B{n}) \sep t \\
& \ELSIF \B{n} \mbox{ is a local variable } \THEN 
\llvm{""} \sep \local(\B{n}) \sep t \\
& \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \name \AND v = \name \IN \\
& \quad \llvm{"\PH{p} = getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
\end{align*}

\paragraph{Comparisons} are formed by the application of a relational operator
to two expressions. Mapping $\opmap$ associates B operators to corresponding
LLVM keywords:
$$\opmap = \{ = \mapsto \llvm{"eq"}, 
  \ne \mapsto \llvm{"ne"}, 
  < \mapsto \llvm{"slt"}, 
  \le \mapsto \llvm{"sle"}, 
  > \mapsto \llvm{"sgt"}, 
  \ge \mapsto \llvm{"sge"} \}.
$$

The code responsible for evaluating a comparison \B{r} is defined by the
following rule. First both arguments are evaluated in $p_1$ and $p_2$, and then
an instance of the LLVM comparison instruction \llvm{icmp} is used to compare
the temporary variables $v_1$ and $v_2$, of type $t_1 = t_2$, storing the result
in the fresh variable $v$.
\begin{align*}
\lefteqn{\trad{r}{\B{Comp}} \isdef
  \LET
  p_1 \sep v_1 \sep t_1 = \trad{r.arg1}{\B{Expr}} \AND
  p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
& \AND v = \name \IN \\
& \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = icmp \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v
\end{align*}

\paragraph{Arithmetic operations} have a similar structure and interpretation
as comparisons and their presentation is mostly omitted. Let us present the
translation of the unary \B{succ} operator though:
\begin{align*}
  \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t =
  \trad{n.arg}{\B{expr}} \AND w = \name \IN \\
  & \llvm{"\PH{p} \quad \PH{w} = i32 add 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
\end{align*}

\paragraph{Predicates}
The translation of conditions takes as input a condition node \B{n}, as well as
two labels $\ell_t$ and $\ell_f$ that correspond to program locations where the
execution shall jump when the condition is evaluted to true or false,
respectively.  The first rule defines the translation of an atomic relation:
\begin{align*}
  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{Form}} \isdef
  \textbf{ let } p \sep v = \trad{n}{\B{Comp}} \IN} \\
  & \llvm{"\PH{p} \quad br i1 \PH{v}, label \PH{\ell_t}, label \PH{\ell_f} \nl"}
\end{align*}

\noindent The translation for negation is simply: $\trad{not f, $\ell_t$,
  $\ell_f$}{\B{not}} \isdef \trad{f, $\ell_f$, $\ell_t$}{\B{Form}}$.  The
following rule handles the case of conjunctions. Note that it produces code with
a ``short-cut'' when the evaluation of the first argument yields false:
\begin{align*}
\begin{split}
  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{and}} \isdef \LET \ell = \name \IN} \\
  & \LET p_1 = \trad{n.arg1, $\ell$, $\ell_f$}{\B{Form}}
  \AND p_2 = \trad{n.arg2, $\ell_t$, $\ell_f$}{\B{Form}} \IN \\
  & \quad \llvm{"\PH{p_1} \PH{\ell} :  \nl \PH{p_2}"}
\end{split}
\end{align*}
The definition of the translation for disjunctions is similar and is omitted.

%\section{Translation: Control Flow and Instructions
%  \label{sec:control}}

@@

%This section addresses how operations in B implementations are translated to
%LLVM functions. It also discusses initialization, which might be seen as a
%special kind of operation. For operations, the presentation of the translation
%is divided into three steps: the signature, the local data, and the body
%instructions.

@@

%The signature of the function is composed of the name of the function, the
%result type, the name and type of the parameters. The signature may carry other
%information such as linkage, visibility, calling convention, etc. The B parser
%provides this information after the type checking phase has been completed.  On
%the one hand LLVM functions may only have one result, on the other hand B0
%operations may produce several results. The translation of output parameters
%will be function parameters passed by reference, i.e. they are pointers. For
%uniformity of treatment, the return type of the produced LLVM functions is
%\llvm{void}. Translation of operation headers is defined in
%sec.~\ref{sec:trad-header}. The memory to store local variables is allocated on
%the stack, using the \llvm{alloca} instruction. We describe a function in
%sec.~\ref{sec:trad-alloc} that, applied to a B operation, returns the name and
%type of its local variables used in the operation. These names are translated to
%conform to LLVM restrictions. The body of the operation has to be transformed to
%a sequence of LLVM blocks of statements. For each kind of instruction one (or
%more) translation function is defined. The details are provided in
%section~\ref{sec:trad-instr}.  The results of the translation of each aspect are
%combined as follows:

@@

%  \lefteqn{\trad{\B{op}}{\B{oper}} \isdef} \\
%  & \LET h = \trad{\B{op}}{sig} \AND   a = \trad{\B{op.body}}{alloc} \AND i = \trad{\B{op.body}, \llvm{exit}}{\ListOf{\B{Inst}}}_L \IN  \\
%  & \llvm{"define void \PH{h} \{\nl} \\
%  & \quad \llvm{entry:\nl} \\
%  & \quad \quad \PH{a} \\
%  & \quad \quad \PH{i} \\
%  & \quad \llvm{exit: ret void\nl} \\
%  & \llvm{\}\nl"}

@@

%Initialization of a B implementation shares many characteristics with operations
%but needs to deal with initialization of imported modules. It is discussed in
%sec.~\ref{sec:trad-init}.
%\subsection{Sub-routines signatures}
%\label{sec:trad-header}
%The following rule specifies the translation of operation headers. Its
%parameters are the operation identifier, inputs, outputs, the name of the
%machine it belongs to, and the results are the LLVM function definition
%header: \\
%\noindent$\trad{\B{o}}{sig} \isdef \llvm{"\PH{\Global(\B{o})}(\PH{\state(\B{o.root})}* \%self\$, \PH{\trad{\B{o.inp}}{\ListOf{inp}}}, \PH{\trad{\B{o.out}}{\ListOf{out}}})"}$
%Next are the rules defining the translation of individual input and output
%parameters. They have local scope and they are translated by prepending
%\llvm{\%}:
%  \trad{i}{inp} \isdef 
%  \llvm{"\PH{\trad{i.type}{\B{Type}}} \PH{\local(\B{i})}"} \quad \quad
%  \trad{o}{out} \isdef 
%  \llvm{"\PH{\trad{o.type}{\B{Type}}}* \PH{\local(\B{o})}"}
%\subsection{Stack allocation}
%\label{sec:trad-alloc}
%B operations may have local variables, in which case the translator needs to
%issue LLVM stack allocation instructions. Such instructions have two arguments:
%the type of values to store, and a temporary variable that is assigned the
%address (see figure~\ref{fig:llvm-grammar} for the concrete syntax). The
%generation of stack allocation code visits the parse tree of the operation body
%in a pre-order depth-first traversal, generating one statement for each local
%variable declaration found. The detailed definition of this traversal is omitted
%for space reasons. For each B0 variable \B{v} found, a LLVM name is thus
%created.
%  \trad{v}{alloc} \isdef 
%  \llvm{"\PH{\local(\B{v})} = alloca \PH{\trad{v.type}{\B{Type}}} \nl"}
%After the stack allocations, the LLVM source code contains the statements that
%encodes the instructions in the body of the operation, as described in the
%following section.
%\subsection{Translation of instructions}
%\label{sec:trad-instr}
%Function $\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L$ takes as input a sequence
%of B instructions \B{il} and a label $\ell$ and produces the source code of a
%LLVM block with $\ell$ as exit point. It is defined as follows:
%\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L \isdef & \IF \B{il} \mbox{ is empty} \THEN \llvm{"br \PH{\ell} \nl"} \\
%& \ELSE \LET i, il' = il  \AND p = \trad{il, $\ell$}{\ListOf{\B{Inst}}}_L \IN \\
%& \quad \IF i \mbox{ is \B{If} or \B{While}} \THEN \LET \ell' = \name \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}} \PH{\ell'}: \PH{p}"} \\
%& \quad \ELSE \llvm{"\PH{\trad{i}{\B{Inst}}} \PH{p}"}
%Eventually, $\trad{}{\B{Inst}}$ defines the translation for individual
%instructions. Its definition is specialized for each class of instructions as
%detailed in the rest of this section. Note that a new block is created after
%each \B{If} and \B{While} instruction.
%\paragraph{While instructions} are translated in two blocks of instructions:
%block $\ell_1$ evaluates the loop condition and block $\ell_2$ executes the loop
%body:
%\trad{w, $\ell$}{\B{While}} \isdef 
%  & \LET \ell_1 = \name \AND \ell_2 = \name \AND c \sep v = \trad{w.cond}{\B{Pred}} \IN \\
%  & \llvm{"\PH{\ell_1} : \PH{c}} \\
%  & \llvm{br i1 \PH{v}, label \PH{\ell_2}, label \PH{\ell} \nl} \\
%  & \llvm{\PH{\ell_2} : \PH{\trad{i.body, $\ell_1$}{\ListOf{\B{Inst}}}_L}}

@@

%\paragraph{Becomes equal instructions} must evaluate the target and the source
%of the assignment, and then copy the result of the latter in the former, using
%the LLVM instruction \llvm{store}. The evaluation of the target is defined with
%$\trad{}{lv}$, which yields the corresponding LLVM source code $l$ and the
%variable $p$ containing the assigned location. The evaluation of the source
%yields the corresponding code $r$, the variable $v$ holding the result, and its
%type $t$.
%  \trad{a}{\B{Beq}} \isdef 
%  & \LET l \sep p = \trad{i.lhs}{lv} \AND r \sep v \sep t = \trad{i.rhs}{\B{Expr}} \IN \\
%  & \llvm{"\PH{l}} \\
%  & \llvm{\PH{r}} \\
%  & \llvm{store \PH{t} \PH{v}, \PH{t}* \PH{p} \nl"}
%%   & l \conc r \conc 
%%   \llvm{"store "} \conc t \conc \texttt{" "} \conc v \conc \llvm{", "}
%%   \conc t \conc \llvm{"* "} \conc p \conc \nl

>>>>

LLVM aims to be a general, language agnostic, intermediate representation for
compilers. The compiler front-end translates the source programming language to
LLVM source code, upon which optimization and other static analysis tasks may be
performed, and the compiler back-end translates to the target platform assembly
language. One key feature of LLVM is that it is a single-static assignment (SSA)
language: a variable may only be assigned once. So for each assignment to a
variable found in the source language, the front-end shall generate a new
temporary variable in the LLVM program. For illustration,
figure~\ref{fig:ex-llvm} presents a LLVM program (for clarity, the corresponding
C code appears on the left) with two temporary variables \llvm{\%0} and
\llvm{\%1}.
\begin{figure}
\vspace*{-2.4em}
  \begin{minipage}[t]{.3\textwidth}
    \begin{tabbing}
      foo \= \kill  \\
%      \B{VARIABLES}\\
%      \> y\\
      inc(x) =\\
      \B{PRE} x : INT \B{THEN}   \\
      \> 	y := x+y\\
      \B{END} \\
      \end{tabbing}
  \end{minipage}
  \:\vline\:
  \begin{minipage}[t]{.2\textwidth}
    \begin{tabbing}
      foo \= \kill  \\
%	 int y;\\      
      void inc(int x) \{\\
      \> 	y = x+y; \\
      \}\end{tabbing}
  \end{minipage}
	  \:\vline\:
  \begin{minipage}[t]{.6\textwidth}
    \begin{tabbing}
      foo \= \kill  \\
%	 \llvm{@y = global i32 0 } \\      
      \llvm{define void @inc(i32 \%x) nounwind \{} \\
      \llvm{entry:}                      \\
      \> \llvm{\%0 = load i32* @y, align 4}     \\
      \> \llvm{\%1 = add nsw i32 \%0, \%x}     \\
      \> \llvm{store i32 \%1, i32* @y, align 4} \\
      \> \llvm{ret void} \\
      \llvm{\}}
    \end{tabbing}

@@

%\section{Translation: Expressions
%  \label{sec:expr}}

@@

%To evaluate a composed expression using the LLVM, one must first generate a
%sequence of instructions to evaluate each argument, the resulting value being
%stored in a temporary variable, and second generate an instance of the LLVM
%instruction corresponding to the expression argument.  Translation of
%expressions is defined as functions that take as input some class of expressions
%and produce a triple of strings $p \sep v \sep t$, where $p$ is a possibly empty
%sequence of instructions required in the evaluation (e.g.  for the
%sub-expressions), $v$ is the LLVM expression representing the resulting value,
%and $t$ represents the LLVM type of the expression. In the case of composed
%predicates, though, the evaluation is different as boolean operations need to be
%implemented with conditional and unconditional jumps.

@@

%\paragraph{Atomic expressions} have a simple translation. In the case of
%literals, one only needs to issue the corresponding LLVM literal:
%$\trad{TRUE}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"1"} \sep \llvm{"i1"}$,
%$\trad{FALSE}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"0"} \sep \llvm{"i1"}$ and
%$\trad{l}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"\PH{l}"} \sep \llvm{"i32"}$ when
%$\B{l.type} = \B{INT}$.
%For identifiers, three cases are possible: global constants, local variables,
%and module state variables. For state variables, the translation issues
%statements to assign the address of the corresponding element in the state
%structure to pointer variable $p$ and then dereference $p$ to assign $v$:
%\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
%& \IF \B{n} \mbox{ is a constant } \THEN \llvm{""} \sep \Global(\B{n}) \sep t \\
%& \ELSIF \B{n} \mbox{ is a local variable } \THEN 
%\llvm{""} \sep \local(\B{n}) \sep t \\
%& \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \name \AND v = \name \IN \\
%& \quad \llvm{"\PH{p} = getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
%& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
%\paragraph{Comparisons} are formed by the application of a relational operator
%to two expressions. Mapping $\opmap$ associates B operators to corresponding
%LLVM keywords:
%$$\opmap = \{ = \mapsto \llvm{"eq"}, 
%  \ne \mapsto \llvm{"ne"}, 
%  < \mapsto \llvm{"slt"}, 
%  \le \mapsto \llvm{"sle"}, 
%  > \mapsto \llvm{"sgt"}, 
%  \ge \mapsto \llvm{"sge"} \}.
%$$
%The code responsible for evaluating a comparison \B{r} is defined by the
%following rule. First both arguments are evaluated in $p_1$ and $p_2$, and then
%an instance of the LLVM comparison instruction \llvm{icmp} is used to compare
%the temporary variables $v_1$ and $v_2$, of type $t_1 = t_2$, storing the result
%in the fresh variable $v$.
%\lefteqn{\trad{r}{\B{Comp}} \isdef
%  \LET
%  p_1 \sep v_1 \sep t_1 = \trad{r.arg1}{\B{Expr}} \AND
%  p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
%& \AND v = \name \IN \\
%& \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = icmp \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v
%\paragraph{Arithmetic operations} have a similar structure and interpretation
%as comparisons and their presentation is mostly omitted. Let us present the
%translation of the unary \B{succ} operator though:
%  \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t =
%  \trad{n.arg}{\B{expr}} \AND w = \name \IN \\
%  & \llvm{"\PH{p} \quad \PH{w} = i32 add 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
%\paragraph{Predicates}
%The translation of conditions takes as input a condition node \B{n}, as well as
%two labels $\ell_t$ and $\ell_f$ that correspond to program locations where the
%execution shall jump when the condition is evaluted to true or false,
%respectively.  The first rule defines the translation of an atomic relation:
%  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{Form}} \isdef
%  \textbf{ let } p \sep v = \trad{n}{\B{Comp}} \IN} \\
%  & \llvm{"\PH{p} \quad br i1 \PH{v}, label \PH{\ell_t}, label \PH{\ell_f} \nl"}
%\noindent The translation for negation is simply: $\trad{not f, $\ell_t$,
%  $\ell_f$}{\B{not}} \isdef \trad{f, $\ell_f$, $\ell_t$}{\B{Form}}$.  The
%following rule handles the case of conjunctions. Note that it produces code with
%a ``short-cut'' when the evaluation of the first argument yields false:

@@

%  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{and}} \isdef \LET \ell = \name \IN} \\
%  & \LET p_1 = \trad{n.arg1, $\ell$, $\ell_f$}{\B{Form}}
%  \AND p_2 = \trad{n.arg2, $\ell_t$, $\ell_f$}{\B{Form}} \IN \\
%  & \quad \llvm{"\PH{p_1} \PH{\ell} :  \nl \PH{p_2}"}

@@

%The definition of the translation for disjunctions is similar and is omitted.

\section{Translation: Control Flow and Instructions
  \label{sec:control}}

This section addresses how operations in B implementations are translated to
LLVM functions. It also discusses initialization, which might be seen as a
special kind of operation. For operations, the presentation of the translation
is divided into three steps: the signature, the local data, and the body
instructions.

The signature of the function is composed of the name of the function, the
result type, the name and type of the parameters. The signature may carry other
information such as linkage, visibility, calling convention, etc. The B parser
provides this information after the type checking phase has been completed.  On
the one hand LLVM functions may only have one result, on the other hand B0
operations may produce several results. The translation of output parameters
will be function parameters passed by reference, i.e. they are pointers. For
uniformity of treatment, the return type of the produced LLVM functions is
\llvm{void}. Translation of operation headers is defined in
sec.~\ref{sec:trad-header}. The memory to store local variables is allocated on
the stack, using the \llvm{alloca} instruction. We describe a function in
sec.~\ref{sec:trad-alloc} that, applied to a B operation, returns the name and
type of its local variables used in the operation. These names are translated to
conform to LLVM restrictions. The body of the operation has to be transformed to
a sequence of LLVM blocks of statements. For each kind of instruction one (or
more) translation function is defined. The details are provided in
section~\ref{sec:trad-instr}.  The results of the translation of each aspect are
combined as follows:
\begin{align*}
\begin{split}
  \lefteqn{\trad{\B{op}}{\B{oper}} \isdef} \\
  & \LET h = \trad{\B{op}}{sig} \AND   a = \trad{\B{op.body}}{alloc} \AND i = \trad{\B{op.body}, \llvm{exit}}{\ListOf{\B{Inst}}}_L \IN  \\
  & \llvm{"define void \PH{h} \{\nl} \\
  & \quad \llvm{entry:\nl} \\
  & \quad \quad \PH{a} \\
  & \quad \quad \PH{i} \\
  & \quad \llvm{exit: ret void\nl} \\
  & \llvm{\}\nl"}
\end{split}
\end{align*}

Initialization of a B implementation shares many characteristics with operations
but needs to deal with initialization of imported modules. It is discussed in
sec.~\ref{sec:trad-init}.

%\subsection{Sub-routines signatures}
%\label{sec:trad-header}
%
%The following rule specifies the translation of operation headers. Its
%parameters are the operation identifier, inputs, outputs, the name of the
%machine it belongs to, and the results are the LLVM function definition
%header: \\
%\noindent$\trad{\B{o}}{sig} \isdef \llvm{"\PH{\Global(\B{o})}(\PH{\state(\B{o.root})}* \%self\$, \PH{\trad{\B{o.inp}}{\ListOf{inp}}}, \PH{\trad{\B{o.out}}{\ListOf{out}}})"}$
%
%Next are the rules defining the translation of individual input and output
%parameters. They have local scope and they are translated by prepending
%\llvm{\%}:
%\begin{align*}
%  \trad{i}{inp} \isdef 
%  \llvm{"\PH{\trad{i.type}{\B{Type}}} \PH{\local(\B{i})}"} \quad \quad
%  \trad{o}{out} \isdef 
%  \llvm{"\PH{\trad{o.type}{\B{Type}}}* \PH{\local(\B{o})}"}
%\end{align*}
%
%\subsection{Stack allocation}
%\label{sec:trad-alloc}
%
%B operations may have local variables, in which case the translator needs to
%issue LLVM stack allocation instructions. Such instructions have two arguments:
%the type of values to store, and a temporary variable that is assigned the
%address (see figure~\ref{fig:llvm-grammar} for the concrete syntax). The
%generation of stack allocation code visits the parse tree of the operation body
%in a pre-order depth-first traversal, generating one statement for each local
%variable declaration found. The detailed definition of this traversal is omitted
%for space reasons. For each B0 variable \B{v} found, a LLVM name is thus
%created.
%\begin{align*}
%  \trad{v}{alloc} \isdef 
%  \llvm{"\PH{\local(\B{v})} = alloca \PH{\trad{v.type}{\B{Type}}} \nl"}
%\end{align*}
%
%After the stack allocations, the LLVM source code contains the statements that
%encodes the instructions in the body of the operation, as described in the
%following section.
%
\subsection{Translation of instructions}
\label{sec:trad-instr}

Function $\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L$ takes as input a sequence
of B instructions \B{il} and a label $\ell$ and produces the source code of a
LLVM block with $\ell$ as exit point. It is defined as follows:

\begin{align*}
\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L \isdef & \IF \B{il} \mbox{ is empty} \THEN \llvm{"br \PH{\ell} \nl"} \\
& \ELSE \LET i, il' = il  \AND p = \trad{il, $\ell$}{\ListOf{\B{Inst}}}_L \IN \\
& \quad \IF i \mbox{ is \B{If} or \B{While}} \THEN \LET \ell' = \name \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}} \PH{\ell'}: \PH{p}"} \\
& \quad \ELSE \llvm{"\PH{\trad{i}{\B{Inst}}} \PH{p}"}
\end{align*}

Eventually, $\trad{}{\B{Inst}}$ defines the translation for individual
instructions. Its definition is specialized for each class of instructions as
detailed in the rest of this section. Note that a new block is created after
each \B{If} and \B{While} instruction.

\paragraph{While instructions} are translated in two blocks of instructions:
block $\ell_1$ evaluates the loop condition and block $\ell_2$ executes the loop
body:
\begin{align*}
\begin{split}
\trad{w, $\ell$}{\B{While}} \isdef 
  & \LET \ell_1 = \name \AND \ell_2 = \name \AND c \sep v = \trad{w.cond}{\B{Pred}} \IN \\
  & \llvm{"\PH{\ell_1} : \PH{c}} \\
  & \llvm{br i1 \PH{v}, label \PH{\ell_2}, label \PH{\ell} \nl} \\
  & \llvm{\PH{\ell_2} : \PH{\trad{i.body, $\ell_1$}{\ListOf{\B{Inst}}}_L}}
\end{split}
\end{align*}

@@

\paragraph{Becomes equal instructions} must evaluate the target and the source
of the assignment, and then copy the result of the latter in the former, using
the LLVM instruction \llvm{store}. The evaluation of the target is defined with
$\trad{}{lv}$, which yields the corresponding LLVM source code $l$ and the
variable $p$ containing the assigned location. The evaluation of the source
yields the corresponding code $r$, the variable $v$ holding the result, and its
type $t$.
\begin{align*}
\begin{split}
  \trad{a}{\B{Beq}} \isdef 
  & \LET l \sep p = \trad{i.lhs}{lv} \AND r \sep v \sep t = \trad{i.rhs}{\B{Expr}} \IN \\
  & \llvm{"\PH{l}} \\
  & \llvm{\PH{r}} \\
  & \llvm{store \PH{t} \PH{v}, \PH{t}* \PH{p} \nl"}
%   & l \conc r \conc 
%   \llvm{"store "} \conc t \conc \texttt{" "} \conc v \conc \llvm{", "}
%   \conc t \conc \llvm{"* "} \conc p \conc \nl
\end{split}
\end{align*}
